### 2017-04-05 sections/conclusion.tex
Update on Overleaf.

The implemented algorithm is able to compress arbitrary data using the low-rank approximation matrices. If the max allowed error was set to 1e-5 then the compression ratio was at least 10\% for all tested results. In many cases compression ratio can be even much lower -- bellow 1\% of the original size. The important property of compression algorithm is the fact that the approximation error can be set in advance and there is a guarantee that it will not be exceeded.

The main disadvantage is the computational complexity of the compression algorithm. SVD decomposition is very time consumptive operation. However, this operation is performed only once after FEM analysis is finished and before the post-processing is started. Also, randomized version of SVD decomposition algorithm is much faster if you don't mind the slight increase of approximation error. % da se urychlit pomoci randomized SVD, pokud dopredu znam kompresni pomer a neni pro mne kriticke neprekrocit maximalni chybu.

Other complication is the necessity to use special format to store compressed data in the form of SVD decomposition and the post-processor must be updated to use this new format. Post-processor must perform matrix multiplication to get the original results. However, in usual case the data for one analysis step are needed at a time. In this case only a partial multiplication of single row is enough.

>>>>

The implemented algorithm is able to compress arbitrary data using the low-rank approximation matrices. If the maximum allowed error was set to $10^{-5}$ then the compression ratio was at least 10\% for all tested results. In many cases compression ratio can be even much lower -- bellow 1\% of the original size. The important property of compression algorithm is the fact that the approximation error can be set in advance and there is a guarantee that it will not be exceeded.

The main disadvantage is the computational complexity of the compression algorithm. SVD decomposition is very time consuming operation. However, this operation is performed only once after FEM analysis is finished and before the post-processing is started. Also, randomized version of the SVD decomposition algorithm is much faster and can be used if the slight increase of approximation error is tolerated. % da se urychlit pomoci randomized SVD, pokud dopredu znam kompresni pomer a neni pro mne kriticke neprekrocit maximalni chybu.

Other complication is the necessity to use special format to store compressed data in the form of SVD decomposition and the post-processor must be updated to use this new format. Post-processor must perform matrix multiplication to get the original results. However, in usual case the data for one analysis step are needed at a time, i.e. only a partial multiplication of a single row is enough.

### 2017-04-05 sections/implementation.tex
Update on Overleaf.

Compression method can be lossy or lossless based on quality of data reconstructed of its compressed representation. Lossless methods are able to fully recreate original data. Lossy methods on the other hand produce only approximations of original data.

SVD is used as part of the compression algorithm. SVD method applied to arbitrary matrix produces decomposition that consist of corresponding singular values and singular vectors. This process is fully reversible (with the assumption that the numerical errors are negligible). The original matrix can be reconstructed by the multiplication of decomposed parts. However, the compression algorithm is based on modification of decomposition to create \textbf{low-rank approximation matrix}. The reconstructed matrix slightly differs from the original matrix and algorithm therefore does \textbf{lossy} compression.

SVD decomposition is applied on matrices. The first thing to do is therefore to assemble an input matrix. Results from the Finite element method are discrete values calculated in nodes or integration points. There are usually multiple fields in result data (e.g. Displacements, Stress, Strain etc.) and each field can have multiple components (e.g. Displacements can have $X$, $Y$ and $Z$ component). As non-linear analyses usually lead to multiple computation steps there are multiple sets of data per each field component. All data corresponding to a single field component form the input matrix to the compression algorithm. Let's call it matrix $\mtrx{A}$. Each row of matrix $\mtrx{A}$ corresponds to single time step and each column corresponds to a single node or integration point.

Once the matrix $\mtrx{A}$ is built for a field component the compression algorithm can be applied on it. It is purely algebraic procedure, no information about geometry of the mesh is needed.

@@

Only single field component is considered to fill the input matrix $\mtrx{A}$. Results for different data components (even for the same field) can be independent from each other. There can also be order-of-magnitude difference between the components and the compression algorithm could potentially clear the important information in the weaker component. Another reason against merging of components is computational complexity. The smaller the matrix $\mtrx{A}$ is, the faster the decomposition algorithm performs.

@@

where $S_{i}$ is $i$-th singular value, $\mtrx{U_{i}}$ and $\mtrx{V_{i}}$ are corresponding singular vectors, and $n$ is the rank of matrix~$\mtrx{A}$. Considering the fact that singular values are ordered $S_{1} \geq S_{2} \geq S_{3} \geq ... \geq S_{n}$ the above formula implies that the first term would have the highest contribution and last term would have the lowest contribution to matrix~$\mtrx{A}$. Therefore, if we take only first $r$ members of the above summation we get an approximation matrix

@@

where $m$ and $n$ are dimensions of matrix $\mtrx{A^{m \times n}}$. Explanation of the compression ratio formula is best done using Figure~\ref{fig:lowrank_svd}. Light color represents the part of matrix decomposition that is to be stored in the output file as a low-rank approximation of the input.

@@

Considering the usual shape of matrix containing FEM results this inequality is easily satisfiable even for the $r$ being close to the rank of the original matrix as in the typical case the number of nodes or integration-points is much higher than the number of analysis steps and therefore $m<<n$.

@@

where $CR$ is the compression ratio, $0 \leq CR \leq 1$ ($0$ results in absolute compression while $1$ results in no compression).

@@

where $k=min(m, n)$, i.e. the smallest of two dimensions of the matrix $\mtrx{M}$. The above formula states that the sum of squared elements of matrix $\mtrx{A}$ equals to the sum of squared singular values $\sigma_{i}$ of the same matrix $\mtrx{A}$.

@@

where the term on the right-hand side is the sum of squares of those singular values of the matrix $A$ that are going to be cut away by the compression algorithm. The equation can be rewritten using the definition of $MSE$ in \eqref{eq:mse-def} to

@@

and using \eqref{eq:nrmsd-def} further to

@@

where $e$ is NRMSD used as an error limit that can not be exceeded to achieve reasonable quality of approximation.

\begin{algorithm}
  \caption{Calculation of rank for approximation matrix from max allowed error}\label{rankAlgorithm}
  \label{alg:rank-calculation}
  \begin{algorithmic}[1]
  	\INPUT max allowed error ($e$), array with singular values ($S$), element count ($c$), max element value ($x_{max}$), min element value ($x_{min}$)
    \OUTPUT rank of resulting matrix
    \Procedure{CalculateRank}{$e, S, c, x_{max}, x_{min}$}
      \State $MSE \gets 0$
      \State $NRMSD \gets 0$
      \State $rank \gets S.length$
      \While{$NRMSD < e$}\Comment{repeat until max error is reached}
        \State $MSE \gets MSE + S[rank]/c$ \Comment{calculate MSE for current rank}
        \State $NRMSD \gets \sqrt{MSE} / (x_{max} - x_{min})$ \Comment{normalize error}
        \State $rank \gets rank - 1$ \Comment{decrement rank for next loop}
      \EndWhile
      \State \textbf{return} $rank + 1$ \Comment{Add one to not exceed max allowed error}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

Time complexity of the exact SVD decomposition algorithm is $$O(m^2n),$$
% http://mathoverflow.net/questions/161252/what-is-the-time-complexity-of-truncated-svd

where $m<n$. This theoretical algorithm complexity is confirmed by two benchmarks where the dependency of execution time on varying matrix dimension is shown. The results of the benchmarks are depicted in Figure \ref{fig:ExeTime_rows} and Figure \ref{fig:ExeTime_columns}. Several observations were made from the results:

@@

Storage size of SVD decomposition itself can also be optimized in several ways. E.g. $\mtrx{\Sigma}$ being a diagonal matrix can be stored as single list of singular values $\sigma_{i}$ or can be even multiplied with the matrix of left singular vectors $\mtrx{U}$.

>>>>

Compression method can be lossy or lossless based on quality of data reconstructed of its compressed representation. Lossless methods are able to fully reconstruct original data. Lossy methods on the other hand produce only approximations of original data.

SVD is used as part of the compression algorithm. SVD method applied to arbitrary matrix produces decomposition that consist of corresponding singular values and singular vectors. This process is fully reversible (with the assumption that the numerical errors are negligible). The original matrix can be reconstructed by the multiplication of decomposed parts. However, the compression algorithm is based on modification of decomposition to create low-rank approximation matrix. The reconstructed matrix slightly differs from the original matrix and algorithm therefore does lossy compression.

SVD decomposition is applied on matrices. The first thing to do is therefore to assemble an input matrix. Results from the Finite element method are discrete values calculated in nodes or integration points. There are usually multiple fields in result data (e.g. displacements, stress, strain etc.) and each field can have multiple components (e.g. displacements can have $X$, $Y$ and $Z$ component). As non-linear analyses usually lead to multiple computation steps there are multiple sets of data per each field component. All data corresponding to a single field component form the input matrix $\mtrx{A}$ to the compression algorithm. Only single field component is considered to fill the input matrix $\mtrx{A}$. Results for different data components (even for the same field) can be independent from each other. There can also be order-of-magnitude difference between the components and the compression algorithm could potentially clear the important information in the weaker component. Another reason against merging of components is computational complexity. The smaller the matrix $\mtrx{A}$ is, the faster the decomposition algorithm performs.

Each row of matrix $\mtrx{A}$ corresponds to single time step and each column corresponds to a single node or integration point. Once the matrix $\mtrx{A}$ is built for a field component the compression algorithm can be applied on it. It is purely algebraic procedure, no information about geometry of the mesh is needed.

@@

@@

\noindent
where $S_{i}$ is $i$-th singular value, $\mtrx{U_{i}}$ and $\mtrx{V_{i}}$ are corresponding singular vectors, and $n$ is the rank of matrix~$\mtrx{A}$. Considering the fact that singular values are ordered $S_{1} \geq S_{2} \geq S_{3} \geq ... \geq S_{n}$ the above formula implies that the first term would have the highest contribution and last term would have the lowest contribution to matrix~$\mtrx{A}$. Therefore, if we take only first $r$ members of the above summation we get an approximation matrix

@@

\noindent
where $m$ and $n$ are dimensions of matrix $\mtrx{A}$. Explanation of the compression ratio formula is best done using Figure~\ref{fig:lowrank_svd}. Light color represents the part of matrix decomposition that is to be stored in the output file as a low-rank approximation of the input.

@@

\noindent
Considering the usual shape of matrix containing FEM results this inequality is easily satisfiable even for the $r$ being close to the rank of the original matrix as in the typical case the number of nodes or integration-points is much higher than the number of analysis steps and therefore $m<<n$.

@@

\noindent
where $CR$ is the compression ratio, $0 \leq CR \leq 1$ ($0$ results in absolute compression while $1$ results in no compression).

@@

\noindent
where $k=min(m, n)$, i.e. the smallest of two dimensions of the matrix $\mtrx{M}$. The above formula states that the sum of squared elements of matrix $\mtrx{A}$ equals to the sum of squared singular values $\sigma_{i}$ of the same matrix $\mtrx{A}$.

@@

\noindent
where the term on the right-hand side is the sum of squares of those singular values of the matrix $A$ that are going to be cut away by the compression algorithm. The equation can be rewritten using the definition of $MSE$ in \eqref{eq:mse-def} to

@@

\noindent
and using \eqref{eq:nrmsd-def} further to

@@

\noindent
where $e$ is NRMSD used as an error limit that can not be exceeded to achieve reasonable quality of approximation.

\begin{algorithm}
  \caption{Calculation of rank for approximation matrix from maximum allowed error}\label{rankAlgorithm}
  \label{alg:rank-calculation}
  \begin{algorithmic}[1]
  	\INPUT maximum allowed error ($e$), array with singular values ($S$), element count ($c$), maximum element value ($x_{max}$), minimum element value ($x_{min}$)
    \OUTPUT rank of resulting matrix
    \Procedure{CalculateRank}{$e, S, c, x_{max}, x_{min}$}
      \State $MSE \gets 0$
      \State $NRMSD \gets 0$
      \State $rank \gets S.length$
      \While{$NRMSD < e$}\Comment{repeat until max error is reached}
        \State $MSE \gets MSE + S[rank]/c$ \Comment{calculate MSE for current rank}
        \State $NRMSD \gets \sqrt{MSE} / (x_{max} - x_{min})$ \Comment{normalize error}
        \State $rank \gets rank - 1$ \Comment{decrement rank for next loop}
      \EndWhile
      \State \textbf{return} $rank + 1$ \Comment{Add one to not exceed maximum allowed error}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

% http://mathoverflow.net/questions/161252/what-is-the-time-complexity-of-truncated-svd

Time complexity of the exact SVD decomposition algorithm is $O(m^2n)$, where $m<n$. This theoretical algorithm complexity is confirmed by two benchmarks where the dependency of execution time on varying matrix dimension is shown. The results of the benchmarks are depicted in Figure \ref{fig:ExeTime_rows} and Figure \ref{fig:ExeTime_columns}. Several observations were made from the results:

@@

Storage size of SVD decomposition itself can also be optimized. $\mtrx{\Sigma}$ being a diagonal matrix can be stored as single list of singular values $\sigma_{i}$ or can be even multiplied with the matrix of left singular vectors $\mtrx{U}$.

### 2017-04-05 sections/introduction.tex
Update on Overleaf.

In \cite{Benes2016} is presented one possible direction for research in this area -- the replacement of discrete data produced by finite element solver by simple continuous functions. These approximation functions can be then described be low number of parameters. The focus is on \textit{geometry}. The main goal is to find the areas in domain where the output discrete function has predictable development and can be easily replaced by e.g. linear continuous function. Although this approach has great results for some class of functions it is not applicable to problem in general. For some special cases, such as functions with discontinuities, the method has poor results because approximation error is too high and -- what is more important -- it can not be controlled in advance.

>>>>

In \cite{Benes2016} is presented one possible direction for research in this area -- the replacement of discrete data produced by finite element solver by simple continuous functions. These approximation functions can be then described by a small number of parameters. The focus is on geometry. The main goal is to find the areas in domain where the output discrete function has predictable development and can be easily replaced by e.g. linear continuous function. Although this approach has great results for some class of functions it is not applicable to problem in general. For some special cases, such as functions with discontinuities, the method has poor results because approximation error is too high and -- what is more important -- it cannot be controlled in advance.

### 2017-04-05 sections/math.tex
Update on Overleaf.

Singular value decomposition (SVD) \cite{Baker2013, Kalman2002} is based on a theorem from linear algebra which says that a rectangular matrix $\mtrx{A}$ can
be broken down into the product of three matrices - an orthogonal matrix $\mtrx{U}$, a diagonal
matrix $\mtrx{S}$, and the transpose of an orthogonal matrix $\mtrx{V}$:

\begin{equation}
\mtrx{A_{m \times n}} = \mtrx{U_{m \times m}} \mtrx{S_{m \times n}} \mtrx{V^{T}_{n \times n}},
\label{eq:svd-def}
\end{equation}

where $\mtrx{U^{T}U} = \mtrx{I}$, $\mtrx{V^{T}V} = \mtrx{I}$; the columns of $\mtrx{U}$ are orthonormal eigenvectors of $\mtrx{AA^{T}}$ , the
columns of $\mtrx{V}$ are orthonormal eigenvectors of $\mtrx{A^{T}A}$, and $\mtrx{S}$ (sometimes referred to as $\mtrx{\Sigma}$) is a diagonal matrix containing singular values in descending order which are at the same time the square roots of eigenvalues from $\mtrx{U}$ or $\mtrx{V}$.

@@

This is the basic idea behind SVD: taking a high dimensional, highly variable set of data points and reducing it to a lower dimensional space that exposes the substructure of the original data more clearly and orders it from most variation to the least. What makes SVD practical for data compression applications is that you can simply ignore variation below a particular threshold to massively reduce your data but be assured that the main relationships of interest have been preserved.

@@

The objective of data compression is to represent an input with lesser amount of data. However, the data reconstructed from its compressed representation may or may not be able to recreate the exact copy of original image. A compression technique can be lossy or lossless based on the quality of data it restores.

@@

where $A_{ij}$ represents an element of the original matrix and $A'_{ij}$ represents an element of the
reconstructed matrix of dimension $m \times n$.

@@

where $X_{min}$ and $X_{max}$ are elements of input matrix $\mtrx{A}$ with minimum and maximum value, respectively.

@@

Exact SVD of a $m \times n$ matrix has time complexity $O(min(mn^2, m^2n))$ using the "big-O" notation. Assume the data points are in the columns of $\mtrx{A} \in \mtrx{M_{m,n}}(\mathbb{R})$ where $m \leq n$. Note that $\mtrx{AA^T}$ is the dataset covariance matrix. Then a simple method is to randomly choose $k<m$ columns of $\mtrx{A}$ that form a matrix $\mtrx{B}$. Statistically, the SVD of $\mtrx{BB^T}$ will be close to that of $\mtrx{AA^T}$; thus it suffices to calculate the SVD of $\mtrx{B}$, the complexity of which, is only $O(k^2m)$.

>>>>

Singular value decomposition (SVD) \cite{Baker2013, Kalman2002} is based on a theorem from linear algebra which says that a rectangular matrix $\mtrx{A} \in \mathbb{R}^{m \times n}$ can be decomposed into the product of three matrices - an orthogonal matrix $\mtrx{U} \in \mathbb{R}^{m \times m}$, a diagonal
matrix $\mtrx{S} \in \mathbb{R}^{m \times n}$, and the transpose of an orthogonal matrix $\mtrx{V} \in \mathbb{R}^{n \times n}$:

\begin{equation}
\mtrx{A} = \mtrx{U} \mtrx{S} \mtrx{V}^\mathsf{T},
\label{eq:svd-def}
\end{equation}

\noindent
where $\mtrx{U^\mathsf{T}U} = \mtrx{I}$, $\mtrx{V^\mathsf{T}V} = \mtrx{I}$; the columns of $\mtrx{U}$ are orthonormal eigenvectors of $\mtrx{AA^\mathsf{T}}$, the columns of $\mtrx{V}$ are orthonormal eigenvectors of $\mtrx{A^\mathsf{T}A}$, and $\mtrx{S}$ (sometimes referred to as $\mtrx{\Sigma}$) is a diagonal matrix containing singular values in descending order which are at the same time the square roots of eigenvalues of $\mtrx{U}$ or $\mtrx{V}$.

@@

This is the basic idea behind SVD: taking a high dimensional, highly variable set of data points and reducing it to a lower dimensional space that exposes the substructure of the original data more clearly and orders it from most variation to the least. What makes SVD practical for data compression applications is that variation below a particular threshold can be simply ignored to massively reduce data with assurance that the main relationships of interest have been preserved.

@@

The objective of data compression is to represent an input with smaller amount of data. However, the data reconstructed from its compressed representation may or may not be the exact copy of the original image. A compression technique can be lossy or lossless based on the quality of data it restores.

@@

\noindent
where $A_{ij}$ represents an element of the original matrix and $A'_{ij}$ represents an element of the
reconstructed matrix of dimension $m \times n$.

@@

\noindent
where $X_{min}$ and $X_{max}$ are elements of input matrix $\mtrx{A}$ with minimum and maximum value, respectively.

@@

Exact SVD of a $m \times n$ matrix has time complexity $O(min(mn^2, m^2n))$ using the "big-O" notation. Assume the data points are in the columns of $\mtrx{A}$ where $m \leq n$. Note that $\mtrx{AA^\mathsf{T}}$ is the dataset covariance matrix. Then a simple method is to randomly choose $k<m$ columns of $\mtrx{A}$ that form a matrix $\mtrx{B}$. Statistically, the SVD of $\mtrx{BB^\mathsf{T}}$ will be close to that of $\mtrx{AA^\mathsf{T}}$; thus it suffices to calculate the SVD of $\mtrx{B}$, the complexity of which, is only $O(k^2m)$.

### 2017-04-05 sections/results.tex
Update on Overleaf.

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{figures/chotkova_MaxError}
\caption{Dependence of Normalized Max Error on Compression ratio for Chotkova project results.}
\label{fig:chotkova:MaxError}
\end{figure}

% mechaxisym
Figure \ref{fig:mechaxisym:mesh} contains visualization of results of two-dimensional analysis. There are exactly 232 analysis steps. The resulting data has linear function character with several discontinuities in temporal dimension. There are few time steps in which resulting discrete functions have very different values compared to neighboring time steps. This was supposed to have negative impact on the quality of compression. However, as can be seen in Figure \ref{fig:mechaxisym:NRMSD}, the quality is better than expected. E.g. if the rank of approximation matrix is set to 3 (compared to 232 being the rank of the original matrix) the normalized relative error (NRMSD) does not exceed 1e-5.

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{figures/mechaxisym_screenshot}
\caption{"Mechaxisym" model. Results visualization (Displacement field, X component).}
\label{fig:mechaxisym:mesh}
\end{figure}

>>>>

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{figures/chotkova_MaxError}
\caption{Dependence of Normalized Maximum Error on Compression ratio for Chotkova project results.}
\label{fig:chotkova:MaxError}
\end{figure}

% mechaxisym
Figure \ref{fig:mechaxisym:mesh} contains visualization of results of two-dimensional analysis. There are exactly 232 analysis steps. The resulting data has linear function character with several discontinuities in temporal dimension. There are few time steps in which resulting discrete functions have very different values compared to neighboring time steps. This was supposed to have negative impact on the quality of compression. However, as can be seen in Figure \ref{fig:mechaxisym:NRMSD}, the quality is better than expected; e.g., if the rank of approximation matrix is set to 3 (compared to 232 being the rank of the original matrix) the normalized relative error (NRMSD) does not exceed $10^{-5}$.

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{figures/mechaxisym_screenshot}
\caption{"Mechaxisym" model. Results visualization (displacement field, x component).}
\label{fig:mechaxisym:mesh}
\end{figure}
